Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-22 manuscript.qmd
further edits in the formalism

We represent a patient’s state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables, where each variable corresponds to a lymph node level (LNL):

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's hidden binary random variables may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or clinically detected). Since the state vector $\mathbf{X}[t]$ is $V$-dimensional and binary, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. In addition, each LNL is associated with an observed binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that suspicious lymph nodes were detected that are deemed metastatic.

A hidden Markov model (HMM) is defined by three key components:

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state where all LNLs are still healthy $\boldsymbol{\xi}_0$.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. Consequently, many elements of the transition matrix are zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ represent the probability $t_{vr}$ for inter-nodal progression. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].

@@

Using the introduced quantities, we can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$. For later use, we define at this point the matrices of the dsitributions over all hidden states, given all time-steps $\boldsymbol{\Lambda}$:

@@

At the time of the diagnosis $t_D$, we multiply the result with the observation matrix $\mathbf{B}$. We may then look up the likelihood of a patient presenting with the diagnosis $\mathbf{Z}=\boldsymbol{\zeta}_i$ in the $i$-th entry of the final result. However, the remaining issue is that the value of $t_D$ is unknown, i.e. over how many time-steps the HMM should be evolved. We solve this problem by marginalizing over the time of diagnosis. Different distributions over the diagnosis times can then be choosen based on T-category. For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$. This gives us for example

>>>>

We represent a patient’s state of involvement at an abstract time-step $t$ as a vector of hidden binary random variables, where each component corresponds to a lymph node level (LNL):

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to some form of tumor presence (i.e., occult or macroscopic disease). In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved. $X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.

Based on this, our HMM is fully described by defining the following three quantities:

1. A starting state $\mathbf{X}[t=0]$ at time $t=0$ just before the patient's tumor formed. In our case, this is always the state $\boldsymbol{\xi}_0$ where all LNLs are still healthy.
2. The *transition matrix*
   $$
   \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-trans-matrix}
   where the value at row $i$ and column $j$ represents the probability to transition from state $\boldsymbol{\xi}_i$ to $\boldsymbol{\xi}_j$ during the time-step from $t$ to $t+1$. Note that we prohibit self-healing, meaning that during a transition, no LNL may change their state from $X_v[t]=1$ to $X_v[t+1]=0$. Consequently, many elements of the transition matrix are zero.
3. Lastly, the *observation matrix*
   $$
   \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
   $$ {#eq-obs-matrix}
   where in row $i$ and at column $j$ we find the probability to *observe* a lymphatic involvement pattern $\mathbf{Z} = \boldsymbol{\zeta}_j$, given that the true (but hidden) state of involvement at the time of diagnosis $t_D$ is $\mathbf{X}[t_D] = \boldsymbol{\xi}_i$.

The transition matrix $\mathbf{A}$ is parameterized using a directed acyclic graph (DAG) that represents the underlying lymphatic network. Edges from the primary tumor to an LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Arcs from a LNL $v$ to a LNL $r$ are parameterized with the probability rate $t_{vr}$ for inter-nodal progression. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [@ludwig_modelling_2024].

@@

We can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$. For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered time-steps:

@@

At the time of the diagnosis $0 \leq t_D \leq t_\text{max}$, we multiply the result with the observation matrix $\mathbf{B}$ to get the distribution over all possible diagnoses. However, the time of diagnosis $t_D$ is unknown, i.e. we do not know over how many time-steps to evolve the HMM. We solve this problem by marginalizing over the time of diagnosis, possible allowing the diagnosis at any time-step, albeit with different weights. Different distributions over the diagnosis times can then be choosen based on T-category. For instance, the mean of the time-prior to marginalize over the diagnosis time for early T-category patients $P\left( t_D \mid \text{early} \right)$ may be shifted towards earlier times than the one for advanced T-category patients $P\left( t_D \mid \text{early} \right)$. This gives us for example:
